Cost allocation from KKT

### 2020-07-07 paper/playaround.tex
split paper in two, flow-based approach and constraint matrix approach

% general symbols
\newcommand{\generation}[1][n]{g_{#1,s,t}}
\newcommand{\generationpotential}{\bar{g}_{n,s,t}}
\newcommand{\generationshare}[1][n]{\omega_{#1,s,t}}
\newcommand{\nodalgeneration}[1][n]{g_{#1,t}}
\newcommand{\capacityGeneration}{G_{n,s}}
\newcommand{\capacityGenerationUpper}{\bar{G}_{n,s}}
\newcommand{\capacityGenerationLower}{\ubar{G}_{n,s}}
\newcommand{\capacityFlow}{F_{\ell}}
\newcommand{\capacityFlowUpper}{\bar{F}_{\ell}}
\newcommand{\capacityFlowLower}{\ubar{F}_{\ell}}
\newcommand{\capexGeneration}{c_{n,s}}
\newcommand{\capexFlow}{c_{\ell}}
\newcommand{\opexGeneration}[1][n]{o_{#1,s}}
\newcommand{\demand}[1][n]{d_{#1,a,t}}
\newcommand{\nodaldemand}[1][n]{d_{#1,t}}
\newcommand{\demandshare}[1][n]{\omega_{#1,a,t}}
\newcommand{\utility}{U_{n,a,t}}
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\ptdfEqual}[1][n]{\ptdf[#1]^\circ}
\newcommand{\slackflow}{k_{\ell}}
\newcommand{\slack}[1][n]{k_{#1}}
\newcommand{\slackk}[1][n]{k^*_{#1}}
\newcommand{\Slack}{k_{m,n}}
\newcommand{\Slackk}{k^*_{m,n}}
\newcommand{\mulowergeneration}[1][n]{\ubar{\mu}_{#1,s,t}}
\newcommand{\muuppergeneration}[1][n]{\bar{\mu}_{#1,s,t}}
\newcommand{\mulowerflow}{\ubar{\mu}_{\ell,t}}
\newcommand{\muupperflow}{\bar{\mu}_{\ell,t}}
\newcommand{\lmp}[1][n]{\lambda_{#1,t}}
\newcommand{\flow}{f_{\ell,t}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}
\newcommand{\injection}{p_{n,t}}
\newcommand{\netconsumption}[1][n]{p^{-}_{#1,t}}
\newcommand{\netproduction}[1][n]{p^{+}_{#1,t}}
\newcommand{\selfconsumption}[1][n]{p^{\circ}_{#1,t}}

@@

% allocation quantities
\newcommand{\allocatePeer}[1][(m, s) \rightarrow n]{A_{#1,t}}
\newcommand{\allocateFlow}[1][n]{F_{#1,\ell,t}}
\newcommand{\allocateTransaction}[1][m \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocateCapexGeneration}[1][n]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocateCapexFlow}[1][n]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateOpex}[1][n]{\mathcal{O}_{#1,t}}
\newcommand{\allocateEmissionCost}[1][n]{\mathcal{E}_{#1,t}}

\newcommand{\emission}[1][n]{e_{#1,s}}
\newcommand{\emissionPrice}{\mu_{\text{CO2}}}
\newcommand{\megawatthour}{MWh$_\text{el}$}
\newcommand{\totalcost}{\mathcal{TC}}
\newcommand{\totalOpexGeneration}{\mathcal{O}}
\newcommand{\totalOpexFlow}{\mathcal{O}^F}
\newcommand{\totalCapexGeneration}{\mathcal{C}^G}
\newcommand{\totalCapexFlow}{\mathcal{C}^F}
\newcommand{\totalEmissionCost}{\mathcal{E}}

@@

The total cost $\totalcost$ are composed of
\begin{align}
\totalcost &= \totalOpexGeneration + \totalCapexGeneration +  \totalCapexFlow  \\
&= \sum_{n,t} \left(  \pdv{\totalOpexGeneration}{\nodaldemand} + \pdv{\totalCapexGeneration}{\nodaldemand} + \pdv{\totalCapexFlow}{\nodaldemand} \right) \nodaldemand 
\label{eq:total_cost_terms}
\end{align}
% 
The total OPEX is given by 
\begin{align}
 \totalOpexGeneration& = \sum_{m,s,t} \opexGeneration[m]\,  \generation[m] 
\end{align}
% 
The OPEX payments assigned to node $n$ at time step $t$ is given by 
% 
\begin{align}
\allocateOpex  = \pdv{\totalOpexGeneration}{\nodaldemand} \, \nodaldemand & =  \sum_{m,s,t}\opexGeneration[m] \,  \allocatePeer
\end{align}
% 
with $\allocatePeer$ being defined as the power produced by generator $(m, s)$ and consumed at node $n$
\begin{align}
 \allocatePeer = \pdv{\generation[m]}{\nodaldemand} \, \nodaldemand
\end{align}
% 
The derivative $\partial \generation[m] / \partial \nodaldemand$ is not well-defined but implies a degree of freedom. However it is constraint to allocating exactly the amount of power which is produced by generator $(m,s)$, thus 
\begin{align}
\generation[m] = \sum_n \allocatePeer 
\label{eq:solution_space_slack}
\end{align}
% 
The direct payment of node $n$ to generator $(m,s)$ is then  
% 
\begin{align}
 \allocateOpex[n \rightarrow (m,s)] &= 
\opexGeneration[m] \,  \pdv{\generation[m]}{\nodaldemand} \, \nodaldemand
\label{eq:allocate_opexGeneration_detailed}\\
\end{align}

% \begin{align}
% \allocateCapexGeneration[n \rightarrow (m,s)] &= 
% \muuppergeneration[m] \, \generationshare[m] \, \allocatePeer
% \label{eq:allocate_capexGeneration_detailed} \\
% \allocateCapexFlow[n \rightarrow \ell] &=  
% \left( \muupperflow - \mulowerflow\right) \sum_{m} \allocateTransaction  
% \label{eq:allocate_capexFlow_detailed}
% \end{align}

@@

\begin{align}
 \totalCapexFlow & = \sum_\ell \capexFlow \capacityFlow = \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right)  \flow 
\end{align}

The individual cost terms are 
\begin{align}
\allocateCapexFlow = \pdv{\totalCapexFlow}{\nodaldemand}\nodaldemand & = \pdv{\left( \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right)  \flow \right) }{\nodaldemand} \\
&= \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right) \ptdf 
\end{align}

\begin{align}
 \pdv{\muupperflow}{\nodaldemand} = 0 \;\; ?
\end{align}

\begin{align}
\pdv{\lagrangian}{\demand} & = \pdv{\lagrangian}{\muupperflow} \,  \pdv{\muupperflow} {\nodaldemand} \\
&= \left( \flow - \capacityFlow \right)  \,  \pdv{\muupperflow} {\nodaldemand}
\end{align}

\begin{align}
\notag
 \pdv{\lagrangian}{\demand}
 =&    \sum_{m, s, t} \opexGeneration[m] \Slack  \\
\notag
&+ \sum_{m,t} \lmp[m] \left(\sum_\ell \incidence[m] \, \ptdf  - \Slack + \delta_{m,n}  \right)  \\ 
\notag
&+ \sum_{n,s,t} \muuppergeneration \generationshare \Slack %  - \sum_{n,s,t} \mulowergeneration \generation  
\\
&+ \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right) \ptdf      
\label{eq:full_lagrangian}
\end{align}

\subsubsection*{Production CAPEX}
\begin{align}
 \totalCapexGeneration& = \sum_{n,s} \capexGeneration \capacityGeneration\\
 &= \sum_{n,t} \muuppergeneration  \generation 
\end{align}

and taking the derivative

\begin{align}
\pdv{\totalCapexGeneration}{\nodaldemand} & = \pdv{ \left(  \sum_{m,t} \muuppergeneration[m]  \generation[m] \right) }{\nodaldemand} \\
&= \sum_{m,t} \generationshare  \muuppergeneration \, \Slack
\end{align}

>>>>

% general symbols
\newcommand{\generation}{g_{s,t}}
\newcommand{\generationpotential}{\bar{g}_{s,t}}
\newcommand{\generationshare}{\omega_{s,t}}
\newcommand{\incidenceGenerator}{K_{n,s}}
\newcommand{\nodalgeneration}{\incidenceGenerator \, \generation}
\newcommand{\capacityGeneration}{G_{s}}
\newcommand{\capacityGenerationUpper}{\bar{G}_{s}}
\newcommand{\capacityGenerationLower}{\ubar{G}_{s}}
\newcommand{\capacityFlow}{F_{\ell}}
\newcommand{\capacityFlowUpper}{\bar{F}_{\ell}}
\newcommand{\capacityFlowLower}{\ubar{F}_{\ell}}
\newcommand{\capexGeneration}{c_{s}}
\newcommand{\capexFlow}{c_{\ell}}
\newcommand{\opexGeneration}{o_{s}}
\newcommand{\demand}{d_{n,t}}
\newcommand{\demandshare}{\omega_{a,t}}
\newcommand{\utility}{U_{n,a,t}}
\newcommand{\incidence}[1][n]{K_{#1,\ell}}
\newcommand{\ptdf}[1][n]{H_{\ell,#1}}
\newcommand{\ptdfEqual}[1][n]{\ptdf[#1]^\circ}
\newcommand{\slackflow}{k_{\ell}}
\newcommand{\slack}[1][n]{k_{#1}}
\newcommand{\slackk}[1][n]{k^*_{#1}}
\newcommand{\Slack}{k_{m,n}}
\newcommand{\Slackk}{k^*_{m,n}}
\newcommand{\mulowergeneration}{\ubar{\mu}_{s,t}}
\newcommand{\muuppergeneration}{\bar{\mu}_{s,t}}
\newcommand{\mulowerflow}{\ubar{\mu}_{\ell,t}}
\newcommand{\muupperflow}{\bar{\mu}_{\ell,t}}
\newcommand{\lmp}[1][n]{\lambda_{#1,t}}
\newcommand{\flow}{f_{\ell,t}}
\newcommand{\cycle}{C_{\ell,c}}
\newcommand{\impedance}{x_\ell}
\newcommand{\cycleprice}{\lambda_{c,t}}
\newcommand{\injection}{p_{n,t}}
\newcommand{\netconsumption}[1][n]{p^{-}_{#1,t}}
\newcommand{\netproduction}[1][n]{p^{+}_{#1,t}}
\newcommand{\selfconsumption}[1][n]{p^{\circ}_{#1,t}}

@@

% allocation quantities
\newcommand{\allocatePeer}[1][s \rightarrow n]{A_{#1,t}}
\newcommand{\allocateFlow}[1][n]{F_{#1,\ell,t}}
\newcommand{\allocateTransaction}[1][s \rightarrow n]{A_{#1,\ell,t}}
\newcommand{\allocateCapexGeneration}[1][n]{\mathcal{C}^{G}_{#1,t}}
\newcommand{\allocateCapexFlow}[1][n]{\mathcal{C}^{F}_{#1,t}}
\newcommand{\allocateOpex}[1][n]{\mathcal{O}_{#1,t}}
\newcommand{\allocateEmissionCost}[1][n]{\mathcal{E}_{#1,t}}

\newcommand{\emission}[1][n]{e_{#1,s}}
\newcommand{\emissionPrice}{\mu_{\text{CO2}}}
\newcommand{\megawatthour}{MWh$_\text{el}$}
\newcommand{\totalcost}{\mathcal{TC}}
\newcommand{\totalOpexGeneration}{\mathcal{O}}
\newcommand{\totalOpexFlow}{\mathcal{O}^F}
\newcommand{\totalCapexGeneration}{\mathcal{C}^G}
\newcommand{\totalCapexFlow}{\mathcal{C}^F}
\newcommand{\totalEmissionCost}{\mathcal{E}}
\newcommand{\totalRest}{\mathcal{R}}
\newcommand{\totalDemand}{\mathcal{D}}

@@

\section{Optimization}

\def\l{\lambda}
\def\K{\kappa}
\def\m{\mu}
\def\G{\Gamma}
\def\d{\partial}
\def\cL{\mathcal{L}}

Start with a generic linear objective function over $N$ variables $x_i$
\begin{equation}
 \min_{x_i} f(x) =  \min_{x_i}  \sum_{i=1}^N c_i x_i
\end{equation}
such that they respect linear inequality constraints
\begin{equation}
  \sum_i A_{ji} x_i \leq d_j \hspace{1cm} \m_j \hspace{1cm} j=1,\dots M
\end{equation}
    [Linear equality constraints $\sum_i b_i x_i = c$ with $\lambda$ can be replaced by two inequalities $\leq c$, $\geq c$ with $\lambda = \bar{\m} - \ubar{\m}$.]

A subset $B$ of the inequality constraints will be binding at the
optimum point $x^*$, i.e. for $j\in B$, $\sum_i A_{ji} x^*_i = d_j$. We
write $A'_{ji}$ for the matrix that only runs over $j\in B$. For
non-degenerate solutions these will be enough binding constraints to
solve for a unique optimum $x^*$, i.e. $|B| = N \leq M$, so that $A'$ is an $N\times N$
square matrix.  Therefore we can invert the saturation equation to get
\begin{equation}
  x^*_i = \sum_{k\in B} A^{\prime -1}_{ik} d_k
\end{equation}
It might seem that we've now lost all information about the objective function in this expression for $x^*$, but remember that it is the objective function which determines which constraints are binding, i.e. which vertex of the feasible simplex is the optimum.

\section{Power system}

We consider a power system with minimized total cost $\totalcost$ being subject to supplying demand $\demand$ at node $n$ at time $t$. 
The change in total cost against variation of the nodal demand
\begin{align}
\lmp  =  \pdv{\totalcost}{\demand}
\end{align}
defines the Locational Marginal Price $\lmp$ that consumers at node $n$ and time $t$. 
The total cost $\totalcost$ splits into two cost terms
\begin{align}
\totalcost &= \totalDemand +  \totalRest
\label{eq:total_cost_terms}
\end{align}
where $\totalDemand = \sum_i \totalDemand_i $ combines cost all terms which are sensitive to changes in the nodal demand 
\begin{align}
 \pdv{\totalDemand_i}{\demand} \ne 0
\end{align}
and thus determine the LMP,
\begin{align}
\lmp = \pdv{\totalDemand}{\demand} 
\end{align}
and $\totalRest = \sum_i \totalRest_i$ combines cost terms which is not, thus 
\begin{align}
 0 = \pdv{\totalRest_i}{\demand} \Forall{i}
\end{align}
Due to strong duality the total cost $\totalcost$ minus the exogeous cost term $\totalRest$ is totally payed back by the consumers
\begin{align}
 \totalcost - \totalRest = \sum_{n,t} \lmp \demand
\end{align}

\subsubsection*{Allocate OPEX}

The total OPEX is given by 
\begin{align}
 \totalOpexGeneration& = \sum_{s,t} \opexGeneration\,  \generation 
\label{eq:opex_generation}
 \end{align}
 We want to disaggregate the totel OPEX into contribution of bus n at time 
 \begin{align}
  \totalOpexGeneration = \sum_{n,t} \allocateOpex
 \end{align}
which are defined as 
\begin{align}
\allocateOpex  = \pdv{\totalOpexGeneration}{\demand} \, \demand & =  \sum_{s,t}\opexGeneration \, \pdv{\generation}{\demand} \, \demand
\end{align}
% 
From feeding this back into \cref{eq:opex_generation} we see that the derivative $\partial \generation / \partial \demand$ has to fulfill
\begin{align}
\generation = \sum_n \pdv{\generation}{\demand} \, \demand
\end{align}
Where the term at the right hand side denote the power produced by generator $s$ and consumed at node $n$, which we define as
\begin{align}
 \allocatePeer = \pdv{\generation}{\demand} \, \demand
 \label{eq:allocate_peer}
\end{align}
% 
The direct payment of node $n$ to generator $s$ is then  
% 
\begin{align}
 \allocateOpex[n \rightarrow s] &= 
\opexGeneration \,  \allocatePeer
\label{eq:allocate_opexGeneration_detailed}\\
\end{align}

\subsubsection*{Allocate Generation CAPEX}
\begin{align}
 \totalCapexGeneration& = \sum_{s} \capexGeneration \capacityGeneration 
\end{align}
with 
\begin{align}
 \capexGeneration = \muuppergeneration \, \generation
\end{align}
and \cref{eq:allocate_peer}, we define 
\begin{align}
\allocateCapexGeneration = \pdv{\totalCapexGeneration}{\demand} \, \demand & = \sum_{s} \muuppergeneration \, \allocatePeer
\end{align}

\subsubsection*{Allocate Transmission CAPEX}

@@

\begin{align}
 \totalCapexFlow & = \sum_\ell \capexFlow \capacityFlow = \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right)  \flow 
\end{align}
% 
The individual cost terms are 
\begin{align}
\allocateCapexFlow = \pdv{\totalCapexFlow}{\demand}\demand & =  \sum_{\ell,t} \left( \muupperflow - \mulowerflow \right) \pdv{\flow}{\demand} \, \demand
\end{align}
where we assume $\partial \muupperflow/\partial \demand \cong 0 $, as dual variables have a second order dependeny on variables. The marginal change of flow with repect to the nodal demand is again non-unique and requires assumption on where the power is resulting from.

\begin{align}
 \pdv{\flow}{\demand}\, \demand = \pdv{\flow}{\generation} \pdv{\generation}{\demand} \, \demand  = \allocateTransaction
\end{align}
fulfilling 
\begin{align}
\sum_{s,n} \allocateTransaction = \flow 
\end{align}
